Pattern matching for\\ $(231,213)$-avoiding permutations

### 2015-02-02 article.tex
Section 3

	We focus in this paper on $(213,231)$-avoiding permutation. 
	The number of $n$-permutations that avoid both 
	$231$ and $213$ is 
	$t_0 = 1$ for $n = 0$ and 
	$t_n =2^{n−1}$ for $n\geq 1$ \cite{Simion:Schmidt:EJC:1985}.
	The permutations that do not contain the permutation pattern $231$
	are exactly the \emph{stack-sortable permutations} and they are counted by 
	the Catalan numbers \cite{Knuth:1997:ACP:260999}.
	A stack-sortable permutation is a permutation whose elements may be sorted by 
	an algorithm whose internal storage is limited to a single stack data structure.

@@

A \emph{bivincular pattern} (abbreviated BVP) $\sigma$
of length $k$ is a permutation in $S_k$ written in
two-line notation 
(that is the top row is $12\,\ldots\,k$ and the bottom row
is a permutation $\sigma_1\sigma_2\,\ldots\,\sigma_k$).
We have the following conditions on the top and bottom rows
of $\sigma$:
\begin{itemize}
	\item
	If the bottom line of $\sigma$ contains
	$\underline{\sigma_i\sigma_{i+1}\,\ldots\,\sigma_j}$
	then the letters corresponding to 
	$\sigma_i\sigma_{i+1}\,\ldots\,\sigma_j$ in an occurrence of
	$\sigma$ in a permutaton must be adjacent, whereas there is
	no adjacency condition for
	non-underlined consecutive letters.
	Moreover if the bottom row of $\sigma$ begins with
	$_\llcorner{\sigma_1}$ then any occurrence of $\sigma$ 
	in a permutation $\pi$ must begin with the leftmost 
	letter of $\pi$,
	and 
	if the bottom row of $\sigma$ begins with
	${\sigma_k}_\lrcorner$ then any occurrence of $\sigma$ 
	in a permutation $\pi$ must end with the rightmost 
	letter of $\pi$.
	\item
	If the top line of $\sigma$ contains
	$\overline{i\,i+1\,\ldots\,j}$ then the letters corresponding to
	$\sigma_i, \sigma_{i+1}, \ldots, \sigma_j$ in an
	occurrence of $\sigma$ in a permutation must be adjacent in values,
	whereas there is no value adjacency restriction for non-overlined 
	letters.
	Moreover, if the top row of $\sigma$ begins with
	$^ulcorner{1}$ then
	any occurrence of $\sigma$ is a permutation $\pi$ must begin with
	the smallest letter of $\pi$, and
	if top row of $\sigma$ ends with $k^\urcorner$ then
	any occurrence of $\sigma$ is a permutation $\pi$ must end with
	the largest letter of $\pi$.
\end{itemize}

	\section{Pattern Matching Problem With Text and Motif Avoiding (231,213)}

		In this section we focus 
		on the problem of pattern matching if
		both the motif and the text
		avoid (231,213).
		We first exhibit some properties of a stripe 
		of a permutation avoiding (231,213).\\

@@

		\begin{proof} 
		Any other initial element would serve as a `$2$' in either a 
		$231$ or $213$ with $1$ and $n$ as the `$1$' and `$3$' respectively.
		\end{proof}

@@

		\textbf{Lemma \numl{updown}  \thenum} Let $\pi \in \Av{n}{213,231}$,
		$\forall i,j \in [n]$, 
		$i<j$, 
		if $\stripe{\pi}{i}=\ustep$
		and $\stripe{\pi}{j}=\dstep$
		then $\pi(i)<\pi(j)$.\\

		In other word, for every index $i$ and $j$ of an (231,213) avoiding permutation $\pi$,
		if $i$ is an ascent and
		if $j$ is a descent
		then $\pi(i)<\pi(j)$.\\

		\textbf{Proof} By induction on the size of the permutation :
		the proposition is true for the permutations $(1,2)$ and $(2,1)$.
		Suppose the proposition true for $\Av{n}{213,231}$.
		Let $\pi \in \Av{n}{213,231}$, 
		we show that the proposition holds for $\pi$.
		Let $\pi'$ be the permutation $\pi[2:]$.
		$\pi' \in \Av{n}{213,231}$, so
		every element of $p'$ verify our lemma (by induction).
		We still need to compare the first element with the rest.
		But the first element
		is the maximal (minimal) element of all the permutation 
		(by Lemma \ref{lemma:first element is 1 or n}).
		Moreover if the first element is the maximal (minimal)
		it will mapped to an descent (ascent).
		So the proposition is also true
		for the first element.
		$\Box$\\

		\textbf{Lemma \numl{upup} \thenum} Let $\pi \in \Av{n}{213,231}$,
		$\forall i,j \in [n]$, 
		$i<j$, 
		if $\stripe{\pi}{i}=\ustep$
		and $\stripe{\pi}{j}=\ustep$
		then $\pi(i)<\pi(j)$.\\

@@

		\textbf{Lemma \numl{downdown} \thenum}  Let $\pi \in \Av{n}{213,231}$,
		$\forall i,j \in [n]$,
		$i<j$,
		if $\stripe{\pi}{i}=\dstep$
		and $\stripe{\pi}{j}=\dstep$
		then $\pi(i)>\pi(j)$.\\

@@

		\textbf{Proposition \numl{patternMatchingStripe} \thenum} Let $\ptext, \pmotif \in \Av{n}{213,231}$.
		$\pmotif$ is a motif of $\ptext$ if and only if the stripe of $\pmotif$ is a motif of the stripe of $\ptext$.\\

		\textbf{Proof} \\
		$\Rightarrow$ Suppose that $\pmotif$ is a motif of $\ptext$
		by subsequence $s$.
		By definition $s$ normalized is equal to
		$\pmotif$ so they have the same stripe.\\ 
		$\Leftarrow$ 
		Let $\ptext$ be a permutation avoiding $(213,231)$, 
		$S_1$ be its stripe, 
		$\pmotif$  be a permutation avoiding $(213,231)$, 
		and $S_2$ its stripe.
		Let $\phi$ be the function
		that match $S_2$ to $S_1$.
		We define $\varphi$ as, 
		$\forall x \in [|\pmotif|-1]$,
		$\varphi(x) = \phi(x)$,
		and $\varphi(|\pmotif|) = \phi(|\pmotif|-1)+1$
		Suppose that $\exists x,y \in [|\pmotif|]$, 
		$\pmotif(x)<\pmotif(y)$ 
		and $\ptext(\varphi(x))>\ptext(\varphi(y))$,
		without lost of generality we can suppose that $x<y$,
		Let see all the case possible that can make this 
		hypothesis possible :
		\begin{itemize}		
			\item if $\pmotif(x)<\pmotif(y)$, 
			$\stripe{\pmotif}{x}=\ustep$ 
			and $\stripe{\pmotif}{y}=\ustep$
			(if x and y are increasing in $\pmotif$),
			then $\stripe{\ptext}{\varphi(x)} =\ustep$ and $\stripe{\ptext}{\varphi(y)}=\ustep$,
			(the match are also increasing).
			As $x<y$ and as $\varphi$ is an increasing function, 
			$\varphi(x) < \varphi(y)$.
			And thus, by \textbf{Lemma \ref{upup}},
			$\ptext(\varphi(x)) < \ptext(\varphi(y))$ .
			Which contradict our hypothesis,
			so this case is not possible.

			\item if $\pmotif(x)<\pmotif(y)$, 
			$\stripe{\pmotif}{x}=\ustep$ 
			and $\stripe{\pmotif}{y}=\dstep$
			then $\stripe{\ptext}{\varphi(x)}=\ustep$ and $\stripe{\ptext}{\varphi(y)}=\dstep$,
			As $x<y$ and as $\varphi$ is an increasing function, 
			$\varphi(x) < \varphi(y)$.
			And thus, by \textbf{Lemma \ref{updown}},
			$\ptext(\varphi(x)) < \ptext(\varphi(y))$ .
			Which contradict our hypothesis,
			so this case is not possible.

			\item if $\pmotif(x)>\pmotif(y)$, 
			$\stripe{\pmotif}{x}=\dstep$ 
			and $\stripe{\pmotif}{y}=\dstep$
			then $\stripe{\ptext}{\varphi(x)}=\dstep$ and $\stripe{\ptext}{\varphi(y)}=\dstep$,
			As $x<y$ and as $\varphi$ is an increasing function, 
			$\varphi(y) < \varphi(x)$.
			And thus, by \textbf{Lemma \ref{downdown}},
			$\ptext(\varphi(y)) > \ptext(\varphi(x))$ .
			Which contradict our hypothesis,
			so this case is not possible.

			\item if $\pmotif(x)>\pmotif(y)$, 
			$\stripe{\pmotif}{x}=\dstep$ 
			and $\stripe{\pmotif}{y}=\ustep$
			then $\stripe{\ptext}{\varphi(x)}=\dstep$ and $\stripe{\ptext}{\varphi(y)}=\ustep$,
			As $x<y$ and as $\varphi$ is an increasing function, 
			$\varphi(y) < \varphi(x)$.
			And thus, by \textbf{Lemma \ref{updown}},
			$\ptext(\varphi(y)) > \ptext(\varphi(x))$ .
			Which contradict our hypothesis,
			so this case is not possible.$\Box$

@@

		\textbf{Proof}	
		To solve the permutation pattern for a permutation in $\Av{n}{213,231}$, 
		we find a find a match of the stripe of the pattern in the stripe of the text.
		This can be done linearly by a greedy algorithm, 
		by matching a step whenever it is possible.
		And thank to the definition of the definition of a stripe in \textbf{Corollary \ref{stripe}},
		we do not have to compute the stripe advance.
		Therefore we have an linear (online) algorithm.$\Box$\\

>>>>

	We focus in this paper on pattern matching issues for
	$(213,231)$-avoiding permutations.
	The number of $n$-permutations that avoid both 
	$231$ and $213$ is 
	$t_0 = 1$ for $n = 0$ and 
	$t_n =2^{n−1}$ for $n\geq 1$ \cite{Simion:Schmidt:EJC:1985}.
	The permutations that do not contain the permutation pattern $231$
	are exactly the \emph{stack-sortable permutations} and they are counted by 
	the Catalan numbers \cite{Knuth:1997:ACP:260999}.
	A stack-sortable permutation is a permutation whose elements may be sorted by 
	an algorithm whose internal storage is limited to a single stack data structure.

@@

A \emph{bivincular pattern} (abbreviated BVP) $\sigma$
of length $k$ is a permutation in $S_k$ written in
two-line notation 
(that is the top row is $12\,\ldots\,k$ and the bottom row
is a permutation $\sigma_1\sigma_2\,\ldots\,\sigma_k$).
We have the following conditions on the top and bottom rows
of $\sigma$:
\begin{itemize}
	\item
	If the bottom line of $\sigma$ contains
	$\underline{\sigma_i\sigma_{i+1}\,\ldots\,\sigma_j}$
	then the letters corresponding to 
	$\sigma_i\sigma_{i+1}\,\ldots\,\sigma_j$ in an occurrence of
	$\sigma$ in a permutaton must be adjacent, whereas there is
	no adjacency condition for
	non-underlined consecutive letters.
	Moreover if the bottom row of $\sigma$ begins with
	$_\llcorner{\sigma_1}$ then any occurrence of $\sigma$ 
	in a permutation $\pi$ must begin with the leftmost 
	letter of $\pi$,
	and 
	if the bottom row of $\sigma$ begins with
	${\sigma_k}_\lrcorner$ then any occurrence of $\sigma$ 
	in a permutation $\pi$ must end with the rightmost 
	letter of $\pi$.
	\item
	If the top line of $\sigma$ contains
	$\overline{i\,i+1\,\ldots\,j}$ then the letters corresponding to
	$\sigma_i, \sigma_{i+1}, \ldots, \sigma_j$ in an
	occurrence of $\sigma$ in a permutation must be adjacent in values,
	whereas there is no value adjacency restriction for non-overlined 
	letters.
	Moreover, if the top row of $\sigma$ begins with
	$^ulcorner{1}$ then
	any occurrence of $\sigma$ is a permutation $\pi$ must begin with
	the smallest letter of $\pi$, and
	if top row of $\sigma$ ends with $k^\urcorner$ then
	any occurrence of $\sigma$ is a permutation $\pi$ must end with
	the largest letter of $\pi$.
\end{itemize}
A bivincular pattern avoids a set of permutations $P$ if its bottom
row $\sigma_1\sigma_2\,\ldots\,\sigma$ avoids every permutation in $P$.

\section{Both $\pi$ and $\sigma$ are $(231,213)$-avoiding}
\label{section:both are (231,213)-avoiding}

This section is devoted to presenting a fast algorithm for deciding
$\sigma \leq \pi$ 
in case both $\pi$ and $\sigma$ are $(231,213)$-avoiding.

@@

\begin{proof} 
Any other initial element would serve as a `$2$' in either a 
$231$ or $213$ with $1$ and $n$ as the `$1$' and `$3$' respectively.
\qed
\end{proof}

@@

We need the following lemmata.

\begin{lemma}
\label{lemma:updown}
Let $\pi \in \AV_n(213,231)$.
For all $1 \leq i < j \leq n$, 
if $\stripe{\pi}{i}=\ustep$
and $\stripe{\pi}{j}=\dstep$
then $\pi(i)<\pi(j)$.
\end{lemma}

\begin{proof} 
By induction on the size of the permutation.
The proposition is certainly true for the $12$ and $21$.
Suppose the proposition holds true for $\AV_n(213,231)$.
Let $\pi \in \AV_{n+1}(213,231)$ and
$\pi'$ be the permutation $\pi[2:]$.
$\pi' \in \AV_n(213,231)$.
By induction hypothesis, we only need to compare the first
element of $\pi'$ to every element in $\pi'$.
According to lemma~\ref{lemma:first element is 1 or n},
the first element of $\pi$ is either $1$ or $n$.
If the first element of $\pi$ is $1$ then $s_\pi(1) = a$
and $\pi(i) < \pi(j)$ for every $2 \leq j \leq n$.
If the first element of $\pi$ is $n$ then
 $s_\pi(1) = d$.
\qed
\end{proof}

		\begin{lemma}
		\label{lemma:upup}
		Let $\pi \in \AV_n(213,231)$,
		$\forall i,j \in [n]$, 
		$i<j$, 
		if $\stripe{\pi}{i}=\ustep$
		and $\stripe{\pi}{j}=\ustep$
		then $\pi(i)<\pi(j)$.
		\end{lemma}

@@

		\begin{lemma}
		\label{lemma:downdown}
 		Let $\pi \in \AV_n(213,231)$,
		$\forall i,j \in [n]$,
		$i<j$,
		if $\stripe{\pi}{i}=\dstep$
		and $\stripe{\pi}{j}=\dstep$
		then $\pi(i)>\pi(j)$.
		\end{lemma}

@@

		\begin{proposition}
		\label{proposition:patternMatchingStripe}
		Let $\ptext, \pmotif \in \AV_n(213,231)$.
		$\pmotif$ is a motif of $\ptext$ if and only if the stripe of $\pmotif$ is a motif of the stripe of $\ptext$.
		\end{proposition}

		\textbf{Proof} \\
		$\Rightarrow$ Suppose that $\pmotif$ is a motif of $\ptext$
		by subsequence $s$.
		By definition $s$ normalized is equal to
		$\pmotif$ so they have the same stripe.\\ 
		$\Leftarrow$ 
		Let $\ptext$ be a permutation avoiding $(213,231)$, 
		$S_1$ be its stripe, 
		$\pmotif$  be a permutation avoiding $(213,231)$, 
		and $S_2$ its stripe.
		Let $\phi$ be the function
		that match $S_2$ to $S_1$.
		We define $\varphi$ as, 
		$\forall x \in [|\pmotif|-1]$,
		$\varphi(x) = \phi(x)$,
		and $\varphi(|\pmotif|) = \phi(|\pmotif|-1)+1$
		Suppose that $\exists x,y \in [|\pmotif|]$, 
		$\pmotif(x)<\pmotif(y)$ 
		and $\ptext(\varphi(x))>\ptext(\varphi(y))$,
		without lost of generality we can suppose that $x<y$,
		Let see all the case possible that can make this 
		hypothesis possible :
		\begin{itemize}		
			\item if $\pmotif(x)<\pmotif(y)$, 
			$\stripe{\pmotif}{x}=\ustep$ 
			and $\stripe{\pmotif}{y}=\ustep$
			(if x and y are increasing in $\pmotif$),
			then $\stripe{\ptext}{\varphi(x)} =\ustep$ and $\stripe{\ptext}{\varphi(y)}=\ustep$,
			(the match are also increasing).
			As $x<y$ and as $\varphi$ is an increasing function, 
			$\varphi(x) < \varphi(y)$.
			And thus, by \textbf{Lemma \ref{lemma:upup}},
			$\ptext(\varphi(x)) < \ptext(\varphi(y))$ .
			Which contradict our hypothesis,
			so this case is not possible.

			\item if $\pmotif(x)<\pmotif(y)$, 
			$\stripe{\pmotif}{x}=\ustep$ 
			and $\stripe{\pmotif}{y}=\dstep$
			then $\stripe{\ptext}{\varphi(x)}=\ustep$ and $\stripe{\ptext}{\varphi(y)}=\dstep$,
			As $x<y$ and as $\varphi$ is an increasing function, 
			$\varphi(x) < \varphi(y)$.
			And thus, by \textbf{Lemma \ref{lemma:updown}},
			$\ptext(\varphi(x)) < \ptext(\varphi(y))$ .
			Which contradict our hypothesis,
			so this case is not possible.

			\item if $\pmotif(x)>\pmotif(y)$, 
			$\stripe{\pmotif}{x}=\dstep$ 
			and $\stripe{\pmotif}{y}=\dstep$
			then $\stripe{\ptext}{\varphi(x)}=\dstep$ and $\stripe{\ptext}{\varphi(y)}=\dstep$,
			As $x<y$ and as $\varphi$ is an increasing function, 
			$\varphi(y) < \varphi(x)$.
			And thus, by \textbf{Lemma \ref{lemma:downdown}},
			$\ptext(\varphi(y)) > \ptext(\varphi(x))$ .
			Which contradict our hypothesis,
			so this case is not possible.

			\item if $\pmotif(x)>\pmotif(y)$, 
			$\stripe{\pmotif}{x}=\dstep$ 
			and $\stripe{\pmotif}{y}=\ustep$
			then $\stripe{\ptext}{\varphi(x)}=\dstep$ and $\stripe{\ptext}{\varphi(y)}=\ustep$,
			As $x<y$ and as $\varphi$ is an increasing function, 
			$\varphi(y) < \varphi(x)$.
			And thus, by \textbf{Lemma \ref{lemma:updown}},
			$\ptext(\varphi(y)) > \ptext(\varphi(x))$ .
			Which contradict our hypothesis,
			so this case is not possible.$\Box$

@@

		\textbf{Proof}	
		To solve the permutation pattern for a permutation in $\AV_n(213,231)$, 
		we find a find a match of the stripe of the pattern in the stripe of the text.
		This can be done linearly by a greedy algorithm, 
		by matching a step whenever it is possible.
		And thank to the definition of the definition of a stripe in \textbf{Corollary \ref{corollary:stripe}},
		we do not have to compute the stripe advance.
		Therefore we have an linear (online) algorithm.$\Box$\\
